---
title: "Analysis Report 2: Your Title Here"
author: "Emma Gibson"
date: "November 21, 2017"
output: github_document
bibliography: references.bib
csl: bioinformatics.csl
---

*Overall, a single-spaced page is about 500 words. So if the guidelines say half of a page, think about writing around 250 words. You can use the wordcountaddin in RStudio to track your progress.*

# Introduction

Add about 1 page here. Must cite at least 5 peer reviewed articles.

# Methods

## Sample origin and sequencing

Add about half a page here. In this section instead of first person (I/we), use Li et al. [@li2015rna] and Seo et al. [@seo2012transcriptional], since you'll just be describing what they did, based on the methods in their paper(s). It's important to include this so the reader knows what the study design was for the data you will be presenting.

## Computational

These are the methods that were used to process the sequencing data. Should probably be at least a half of a page. At a very minimum should include citations for biomartr, trimmomatic, and sailfish. Note that these three methods references don't count towards the five references you need to cite in the introduction.

# Results

In addition to a minimum of 4-5 figures/tables (and associated captions), you should include sufficient text in this section to describe what your findings were. Remember that in the results section you just describe what you found, but you don't interpret it - that happens in the discussion.

```{r load-libraries, message = FALSE, echo = FALSE}
# Be sure to install these packages before running this script
# They can be installed either with the intall.packages() function
# or with the 'Packages' pane in RStudio

# load general-use packages
library("dplyr")
library("tidyr")
library("knitr")
library("ggplot2")
library("magrittr")

# this package allows for the easy inclusion of literature citations in our Rmd
# more info here: https://github.com/crsh/citr
# and here:
# http://rmarkdown.rstudio.com/authoring_bibliographies_and_citations.html
library("citr")
```

```{r load-data, message = FALSE, echo = FALSE}
# load the dataset from a compressed binary file
# it gets loaded as an object called "final_table"
# this has 3.4 million rows...so you will need to be thoughtful about
# how you analyze the data so that you don't overwhelm your laptop
load("output/final_compiled_counts/joined_count_data.RData")

# changes all columns names to lower case so Travis doesn't complain
names(final_table) %<>% tolower

# test that it loaded correctly before proceeding
stopifnot(exists("final_table"))
```

```{r make-summary-table, echo = FALSE}
# There are many many genes in this dataset, so we can
# subset it down to just a few here to look for interesting patterns
# in the most highly expressed
top_15 <- final_table %>%
  group_by(gender, genename) %>%
  summarize(mean_count = mean(counts_lengthscaledtpm)) %>%
  arrange(desc(mean_count)) %>%
  head(n = 15)

# then we can use the `kable()` function to make a nicely formatted
# markdown table
top_15 %>%
  kable()
```

**Table 1**: The most highly expressed genes in both genders included *SFTPB* and *EEF1A1*.

```{r make-barplot-of-highly-expressed-genes, echo = FALSE}
# this code uses the same data as above, but use it to make a
# barplot - remember geom_col() is just like
# when you use geom_bar(stat = "identity")
top_15 %>%
  ggplot(aes(x = genename,
             y = mean_count,
             fill = gender)) +
    geom_col(position = "dodge")
```

This figure shows the fifteen most common genes in male and female patients.

```{r make-boxplot-of-highly-expressed-genes, echo = FALSE}
# here we just want to pull out the unique gene names and turn
# them into a vector so we can use it below to make a boxplot
# we use the pull() funtion to get this as a vector, just like
# we did when making histograms several weeks ago
top_genes <- top_15 %>%
  ungroup() %>%
  select(genename) %>%
  unique() %>%
  pull()

# now we need to filter from the full data set again, because
# we don't just want summary data, we want all the data in
# order to make boxplots
final_table %>%
  filter(genename %in% top_genes) %>%
  ggplot(aes(x = genename,
             y = counts_lengthscaledtpm,
             fill = gender)) +
    geom_boxplot() +
    facet_wrap(~smoking_status) +
    xlab("Gene Name") +
    ylab("Scaled read counts per gene") +
    ggtitle("Read counts per gene by gender and smoking status") +
    theme_bw() + # simplifies theme
    theme(axis.text.x = # rotates x axis labels vertically
            element_text(angle = 90,
                         hjust = 1))
```

**Figure x**: Here we show another example figure caption.


```{r gender-and-smoking}
# 
gender_and_smoking_status <- final_table %>%
  group_by(smoking_status, cancer_stage, gender, sample_name_s) %>%
  count() %>%
  arrange(desc(n))

names(g_and_ss_tally) <- c("gender", "smoking_status", "cancer_stage", "n")

with(gender_and_smoking_status, table(gender, smoking_status, cancer_stage)) %>%
  as.data.frame() %>%
  ggplot(aes(x = smoking_status,
             y = Freq,
             fill = gender)) +
  facet_wrap(~cancer_stage) +
  geom_col(position = "dodge") +
  ggtitle("Smoking status of patients with each cancer stage") +
  theme(axis.text.x =
            element_text(angle = 90,
                         hjust = 1))
```

```{r female-gene-table}

top_9_f <- final_table %>%
  filter(gender == "female") %>%
  group_by(genename) %>%
  summarize(mean_count = mean(counts_lengthscaledtpm)) %>%
  arrange(desc(mean_count)) %>%
  head(n = 9)

kable(top_15_f)
```


```{r female-gene-graph}

top_genes_f <- top_9_f %>%
  ungroup() %>%
  select(genename) %>%
  unique() %>%
  pull()

final_table %>%
  filter(genename %in% top_genes_f) %>%
  ggplot(aes(x = genename,
             y = counts_lengthscaledtpm,
             fill = gender)) +
    geom_boxplot() +
    facet_wrap(~smoking_status) +
    xlab("Gene Name") +
    ylab("Scaled read counts per gene") +
    ggtitle("Read counts per gene of genes common to female smokers") +
    theme(axis.text.x =
            element_text(angle = 90,
                         hjust = 1))
```



```{r smoking-in-female-genes}
top_f_in_all <- final_table %>%
  filter(genename %in% top_genes_f) %>%
  group_by(gender, genename, cancer_stage) %>%
  summarize(mean_count = mean(counts_lengthscaledtpm))

top_f_in_all %>%
  ggplot(aes(x = genename,
             y = mean_count,
             fill = gender)) +
  facet_wrap(~cancer_stage) +
  geom_col(position = "dodge") +
  ggtitle("Expression of most common genes in female patients in all patients") +
  theme(axis.text.x =
            element_text(angle = 90,
                         hjust = 1))
```

```{r genes-in-m-f-boxplot}

m_f_differences <- final_table %>%
  group_by(gender, genename) %>%
  summarize(mean_count = mean(counts_lengthscaledtpm)) %>%
  spread(key = gender,
         value = mean_count) %>%
  mutate(gender_diff = abs((female - male)/(female + male))) %>%
  filter(!is.nan(gender_diff)) %>%
  arrange(desc(female + male)) %>%
  filter(gender_diff == 1)


```

# Discussion

-it lookslike females have different smoking status/stage
-see if there are genes more common to females than males
-compare to males of same cancer stage
-are these genes due to female smoking status? Or are they differently expressed in males

Add around 1-2 pages interpreting your results and considering future directions one might take in analyzing these data.

# Sources Cited
